---
title: "Estimation of Latent Behavioral States"
author: "Josh Cullen"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document: 
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Behavior Estimation

Movement parameters (SL/TA/TAA) have already been used to delineate time segments of different behavioral patterns. This document displays the output of the subsequent stage: the clustering of behavioral time segments into behavioral states. While some of the components of the clustering model are flexible (e.g., setting the max number of behaviors) or may be ambiguous (e.g., attribution of behaviors to clusters of time segments), this model relies on the user's experience with and knowledge of their study organism. Given prior knowledge about how the focal species moves within their environment, this *a priori* knowledge can be used to guide the assignment of latent behavioral states.

Output of behavioral assignment to time segments (as well as to the whole time series analyzed) will be visualized using multiple methods and components of the data to get a better sense of how the model has clustered time segments, the time series of behavior changes per individual, and the probability that an organism is in each behavioral state by time segment.


## Identifying behaviors

As stated above, time segments were clustered by similar features of their multinomial (SL, TA) or bernoulli (TAA) distributions. An upper limit of 5 possible behaviors was set when running the model over 1000 iterations. Assuming the Gibbs sampler converged after the burn-in period (500 iterations) as determined by an asymptote of the log likelihood, the remaining 500 samples were purported to come from the posterior distribution. In some cases, all posterior samples were used in the analysis, whereas in others only the maximum *a posteriori* (MAP) value was used to assign behavioral states.

To get a sense of the behaviors identified, plots of the movement parameter distributions (SL/TA/TAA) are shown for each of the five behaviors. These behaviors are named according to likely movement based on these distributions. Labels on the x-axis (*theta1.x*,*theta2.x*,*theta3*) represent the bins for TA, SL, and TAA, respectively.
\hfill\break

```{r load}
library('Rcpp')
library('MCMCpack')
library(dplyr)
library(purrr)
library(tidyr) #for gather function
library(ggplot2)
library(ggnewscale) #for multiple fill scales in ggplot2
library(progress) #for progress bar
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(cowplot)
library(reshape2)
library(stringr)
library(lubridate)

#get functions
sourceCpp('aux1.cpp')
source('gibbs functions.R')
source('gibbs sampler2.R')
source('helper functions.R')

#get data
dat<- read.csv('Snail Kite Gridded Data_behav.csv', header = T, sep = ',')
dat.list<- df.to.list(dat)
obs<- get.summary.stats_behav(dat)
obs.list<- df.to.list(obs)

```


```{r gibb sampler, fig.width=4.5, fig.height=3, fig.align='center'}
set.seed(1)

#basic settings
ngibbs=1000
nclustmax=5



#progress bar
pb <- progress_bar$new(
  format = " iteration (:current/:total) [:bar] :percent [Elapsed: :elapsed, Remaining: :eta]",
  total = ngibbs, clear = FALSE, width= 100)

# Run Gibbs sampler
res=cluster.tsegm.behavior(dat=obs[,-1],a.theta3=a.theta3,b.theta3=b.theta3,psi=psi,
                                gamma1=gamma1,nclustmax=nclustmax,ngibbs=ngibbs)

plot(res$loglikel,type='l', xlab = "Iteration", ylab = "Log Likelihood")


#Find MAP value for assignment of clusters
MAP<- which(res$loglikel==max(res$loglikel))  # iteration 711
tbsp.clust<- res$z[MAP,]

#Create DF of clusters by behavioral time segment
behav.seg<- 1:nrow(obs)
tbsp.clust<- cbind(tbsp.clust,behav.seg) %>% data.frame()

```


```{r sep behaviors}
## Separate by Behavior to Compare Distribs (Using Proportions) Over Time

#convert from raw obs to proportion
y1<- obs[,grep('y1',colnames(obs))]
y1.prop<- apply(y1, 1, function(x) {x/sum(x)}) %>% t()

y2<- obs[,grep('y2',colnames(obs))]
y2.prop<- apply(y2, 1, function(x) {x/sum(x)}) %>% t()

y3<- obs[,grep('y3',colnames(obs))]
n<- apply(y2,1,sum)
y3.prop<- y3/n

obs.prop<- cbind(id = obs$id, y1.prop, y2.prop, y3 = y3.prop) %>% data.frame()
```



```{r thetas}
## Separate Histograms of Thetas (not raw proportions) by Behavior

#take mean of posterior theta estimates (after nburn since log likelihood converged)
theta1.post<- res$theta1[501:ngibbs,] %>% apply(2, mean)
theta2.post<- res$theta2[501:ngibbs,] %>% apply(2, mean)
theta3.post<- res$theta3[501:ngibbs,] %>% apply(2, mean)

#create DF with matching thetas per behavior for all observations
theta1<- matrix(theta1.post, nclustmax, ncol(y1), byrow = F)
theta2<- matrix(theta2.post, nclustmax, ncol(y2), byrow = F)
theta3<- matrix(theta3.post, nclustmax, 1, byrow = F)
colnames(theta1)=paste0('theta1.',1:ncol(y1))
colnames(theta2)=paste0('theta2.',1:ncol(y2))
colnames(theta3)='theta3'

theta<- cbind(tbsp.clust = 1:5, theta1, theta2, theta3) %>% data.frame()
theta$tbsp.clust<- factor(theta$tbsp.clust,
                          labels = c("Transit","Exploratory","Resting","ARS","Resting2"))
theta$tbsp.clust<- factor(theta$tbsp.clust, levels(theta$tbsp.clust)[c(1:2,4:5,3)], ordered = T)
theta<- melt(theta, "tbsp.clust")
theta<- theta %>% mutate(param = ifelse(str_detect(theta$variable, "theta1"), "TA",
                                    ifelse(str_detect(theta$variable, "theta2"), "SL", "TAA")))

ggplot(theta, aes(x = variable, y = value, fill = tbsp.clust)) +
  geom_bar(stat = 'identity') +
  scale_fill_viridis_d(guide = F, direction = -1) +
  labs(x = "Bin", y = "Proportion") +
  theme_bw() +
  theme(axis.title = element_text(size = 16), axis.text.y = element_text(size = 14),
        axis.text.x.bottom = element_text(size = 12),
        strip.text = element_text(size = 13), strip.text.x = element_text(face = "bold"),
        axis.text.x = element_text(angle = 90, size = 6)) +
  facet_grid(param ~ tbsp.clust, scales = "free_y")

```
\hfill\break
\hfill\break
\hfill\break
\hfill\break




## Time series of behavior by ID

Now that the behaviors have been roughly defined, I will now show how this looks over time across all individuals. This will provide a better sense of the transitions between behaviors and how long each ID spends in certain behaviors. These plots will be aligned by first observation as well as by date to give a sense of movement patterns as snail kites get older and how this relates to peak breeding season (March 1 to June 30). Additionally, plots that display the proportion of behaviors assigned to a given time segment (from the posterior of the *z's*) will be shown to provide a measure of the uncertainty associated with the behavioral assignments from the MAP value.

```{r}
## Separate by ID Showing Time Series

#create DF with matching proportional distrib by time segment for all observations
obs.prop.time<- obs.prop %>% slice(rep(1:n(), times = n))


#format data
obs.prop.list<- df.to.list(obs.prop.time) %>% lapply(function(x) x[!(names(x) %in% c("id"))])
obs.prop.list.long<- lapply(obs.prop.list, function(x) {x %>% gather(key, value) %>%
    mutate(time=rep(1:nrow(x), times=ncol(x)))})
obs.prop.list.long<- lapply(obs.prop.list.long, function(x) {mutate_at(x, "key", as.factor)})

time<- lapply(obs.prop.list, function(x) data.frame(time = 1:nrow(x))) %>% map_dfr(`[`)
tbsp.clust.time<- tbsp.clust %>% slice(rep(1:n(), times = n)) %>%
  mutate(time = time[,1]) %>% mutate(id = dat$id)



# All IDs

tbsp.clust.time$tbsp.clust<- factor(tbsp.clust.time$tbsp.clust,
                                    labels = c("Transit","Exploratory","Resting","ARS","Resting2"))
tbsp.clust.time$tbsp.clust<- factor(tbsp.clust.time$tbsp.clust,
                                    levels(tbsp.clust.time$tbsp.clust)[c(1:2,4:5,3)], ordered = T)
tbsp.clust.time$id<- factor(tbsp.clust.time$id)
tbsp.clust.time$date<- dat$ESTtime %>% as_datetime()

#Window of peak breeding (March 1 - June 30)
breed<- data.frame(xmin = as_datetime(c("2016-03-01 00:00:00","2017-03-01 00:00:00",
                                        "2018-03-01 00:00:00","2019-03-01 00:00:00")),
                   xmax = as_datetime(c("2016-06-30 23:59:59","2017-06-30 23:59:59",
                                        "2018-06-30 23:59:59","2019-06-30 23:59:59")),
                   ymin = -Inf, ymax = Inf)



## Show Proportion of Behaviors Associated with Each Time Segment (from posterior)

#sample z's from posterior (after burn-in since log-likelihood converged)
z.post<- res$z[501:ngibbs,]
z.post.count<- matrix(0, ncol(z.post), nclustmax)
tmp<-  apply(z.post, 2, table)
for (i in 1:nrow(z.post.count)) {
  z.post.count[i, as.integer(names(tmp[[i]]))]<- tmp[[i]]
}

#convert from count to proportion by time seg
z.post.prop<- apply(z.post.count, 1, function(x) x/sum(x)) %>% t()
colnames(z.post.prop)<- c("Transit","Exploratory","Resting","ARS","Resting2")

#convert to long form for plotting
z.post.prop.time<- cbind(id = obs$id, z.post.prop) %>% data.frame() %>% slice(rep(1:n(), times = n))
z.post.list<- df.to.list(z.post.prop.time) %>% lapply(function(x) x[!(names(x) %in% c("id"))])
z.post.list.long<- lapply(z.post.list, function(x) {x %>% gather(key, value) %>%
    mutate(time=rep(1:nrow(x), times=ncol(x)))})
z.post.list.long<- lapply(z.post.list.long,
                          function(x) {mutate_at(x, "key", factor)})
z.post.list.long<- lapply(z.post.list.long,
                          function(x) {mutate_at(x, "key", ~factor(.,levels(.)[c(5,2,1,4,3)]))})
```

```{r}
#Aligned by first observation
ggplot() +
  geom_tile(data=tbsp.clust.time, aes(x=time, y=id, fill=tbsp.clust)) +
  scale_fill_viridis_d("Behavior", direction = -1) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +
  labs(x = "Time", y = "ID") +
  theme_bw() +
  theme(axis.title = element_text(size = 18), axis.text = element_text(size = 16))
```

```{r}
#Aligned by date
ggplot() +
  geom_rect(data = breed, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            fill = "grey", alpha = 0.5) +
  geom_tile(data=tbsp.clust.time, aes(x=date, y=id, fill=tbsp.clust)) +
  scale_fill_viridis_d("Behavior", direction = -1) +
  scale_y_discrete(expand = c(0,0)) +
  scale_x_datetime(expand = c(0,0)) +
  labs(x = "Time", y = "ID") +
  theme_bw() +
  theme(axis.title = element_text(size = 18), axis.text = element_text(size = 16))
```


```{r posterior z}
# ID 1

ggplot() +
  geom_area(data=z.post.list.long[[1]], aes(x=time, y=value, fill=as.factor(key))) +
  scale_fill_viridis_d("Behavior", direction = -1) +
  scale_y_continuous(expand = c(0,0)) +
  scale_x_continuous(expand = c(0,0)) +
  labs(x = "Time", y = "Proportion of Behavior") +
  theme_bw() +
  theme(axis.title = element_text(size = 18), axis.text = element_text(size = 16))
```



\pagebreak

## Maps of behavioral assignments by ID

Now that a behavioral state has been assigned to all observations (with a 1 hr time step), these values will be plotted in geographic space to visualize how well these classifications match up with perceived movement patterns.
\hfill\break


```{r, fig.align='center', fig.height=8, fig.width=6}
## Generate Geographic Maps with Pts Assigned a Behavior (from MAP estimate)

#load map data
usa <- ne_states(country = "United States of America", returnclass = "sf")
fl<- usa %>% filter(name == "Florida")
fl<- st_transform(fl, crs = "+init=epsg:32617") #change projection to UTM 17N

# tbsp.clust.time$tbsp.clust<- as.factor(tbsp.clust.time$tbsp.clust)
# levels(tbsp.clust.time$tbsp.clust)<- c("Transit","Exploratory","Resting","ARS","Resting2")

dat.list$`1`$behav<- tbsp.clust.time[tbsp.clust.time$id == 1, "tbsp.clust"]
dat.list$`12`$behav<- tbsp.clust.time[tbsp.clust.time$id == 12, "tbsp.clust"]
dat.list$`19`$behav<- tbsp.clust.time[tbsp.clust.time$id == 19, "tbsp.clust"]
dat.list$`27`$behav<- tbsp.clust.time[tbsp.clust.time$id == 27, "tbsp.clust"]
dat<- do.call(rbind.data.frame, dat.list)



# Facet plot of maps

ggplot() +
  geom_sf(data = fl) +
  coord_sf(xlim = c(min(dat$utmlong-120000), max(dat$utmlong+40000)),
           ylim = c(min(dat$utmlat-20000), max(dat$utmlat+20000)), expand = FALSE) +
  geom_path(data = dat, aes(x=utmlong, y=utmlat), color="gray60", size=0.25) +
  geom_point(data = dat, aes(utmlong, utmlat, fill=behav), size=2.5, pch=21, alpha=0.7) +
  scale_fill_viridis_d("Behavior", direction = -1) +
  labs(x = "Longitude", y = "Latitude") +
  theme_bw() +
  facet_wrap(~id)
```



